Investigating magnetic dynamo evolution with TESS field dwarfs

### 2017-10-03 ms.tex
trying to change bib

\newcommand{\columbia}{1}
\newcommand{\simons}{2}
\newcommand{\nyu}{3}
\newcommand{\cca}{4}
\newcommand{\mpia}{5}
\newcommand{\cds}{6}
\newcommand{\naigrain}{333}
\newcommand{\nmcquillan}{100}
\newcommand{\kepexample}{1430163}
\newcommand{\kepexampleperiod}{4}
\newcommand{\aigrainexampleperiod}{20.8}
\newcommand{\Kepler}{{\it Kepler}}
\newcommand{\kepler}{\Kepler}
\newcommand{\corot}{{\it CoRoT}}
\newcommand{\Ktwo}{{\it K2}}
\newcommand{\ktwo}{\Ktwo}
\newcommand{\TESS}{{\it TESS}}
\newcommand{\LSST}{{\it LSST}}
\newcommand{\Wfirst}{{\it Wfirst}}
\newcommand{\SDSS}{{\it SDSS}}
\newcommand{\PLATO}{{\it PLATO}}
\newcommand{\Gaia}{{\it Gaia}}
\newcommand{\gaia}{{\it Gaia}}
\newcommand{\Teff}{$T_{\mathrm{eff}}$}
\newcommand{\teff}{$T_{\mathrm{eff}}$}
\newcommand{\FeH}{[Fe/H]}
\newcommand{\feh}{[Fe/H]}
\newcommand{\ie}{{\it i.e.}}
\newcommand{\eg}{{\it e.g.}}
\newcommand{\logg}{log \emph{g}}
\newcommand{\dnu}{$\Delta \nu$}
\newcommand{\numax}{$\nu_{\mathrm{max}}$}
\newcommand{\acfRMS}{9.27}
\newcommand{\pgramRMS}{1.25}
\newcommand{\mcmcRMS}{0.46}
\newcommand{\racomment}[1]{{\color{red}#1}}

@@

\author{R. Angus\altaffilmark{\columbia, }\altaffilmark{\simons}
        J. R. Davenport, D. Busazi, D. Foreman-Mackey, A. Mann, S. Oh,
        D. Kipping
}

\altaffiltext{\columbia}{Department of Astronomy, Columbia University, NY, NY}
\altaffiltext{\simons}{Simons Fellow, RuthAngus@gmail.com}

@@

% Among the key observable properties of main sequence stars in our Galaxy,
% age is the most difficult to determine.
% Traditionally, fitting isochrones to cluster stars was one of the only precise
% methods for measuring ages but was extremely difficult for the majority of
% isolated field stars, particularly for those without precise spectroscopic
% information.
% Methods such as asteroseismology and measuring Lithium abundances can provide
% precise ages but require time intensive observations for each target and are
% not capable of producing the large quantity of ages needed for exoplanet and
% galactic population studies.

Galactic archaeology and exoplanet populations are two rapidly accelerating
fields of interest within astronomy.
Although seemingly unconnected, these fields are linked by a mutual
requirement for precise stellar parameters.
To galactic archaeologists, ages and elemental abundances are the most
important parameters.
Indeed, most galactic archaeology surveys target exactly these properties.
For exoplaneteers, masses and radii have historically been the most important
stellar parameters for understanding planetary systems.
With a growing number of planet hosts with precise masses and radii, attention
is turning toward other parameters such as ages to understand the history and
evolution of these systems.
Age is therefore a fundamental stellar parameter of great interest to at least
two large communities of astronomers.
However it is a difficult attribute to measure for main sequence F, G, K, and
M stars in the field, in part because low-mass dwarfs do not move far on the
Color-Magnitude diagram (CMD) during their hydrogen burning lifetimes.
Further, competing stellar evolution models predict different ages for the
same star.
Of the measurable properties for a large ensemble of field stars, rotation
periods contain the most information about stellar age, and provide the best
leverage for advancing our knowledge of galactic archeology as well as
exoplanet population demographics via gyrochronology.

Angular momentum is carried away though magnetically driven stellar winds,
which slows the star’s rotation over cosmic time. This rotation-based `clock'
is known as gyrochronology.
Cool spots on the star’s surface rotate in-to and out-of view, creating small
amplitude ($\pm$ 1\%) quasi-periodic changes in the stellar brightness.
While rotation periods have previously been measured from starspot-induced
flux modulations for hundreds of stars from the ground, space-based
photometric surveys have opened the door to homogeneous ensemble measures of
stellar rotation, and therefore age.
Using precise light curves available from the TESS mission, we hope to
infer ages for nearly 100,000 main sequence field stars, including thousands
of planet hosts.
The ages of the planet hosts will allow us to characterize trends in planet
properties over time, by adding to the \kepler\ sample.

Additionally, rotation is related to stellar activity.
With so many TESS targets being M dwarfs which tend to be particularly active,
understanding the magnetic behavior of these stars has never been more
important.

To enable studies of stellar ages from rotation periods with TESS, we propose
to: 1. Measure accurate rotation periods for every available TESS FFI and
two-minute cadence target, using the most appropriate tools and methods
required.
2. Produce updated gyrochronology relations based on a wider range of field
star ages.
3. Determine the origin of the mysterious rotation period bimodality
discovered with Kepler by tracing the rotation period distribution over the
whole sky, and out to further distances utilizing public Gaia data.
4. Measure the star formation history across the sky using a new Bayesian
age-dating system.
5. Further characterize the intrinsic uncertainty in gyrochronology ages using
comoving pairs identified in the first Gaia data release.

@@

The measurement of rotation periods is less sensitive to crowding and source
confusion than exoplanet transit characterization because the rotation period
is not effected by photometric dilution.
We will apply two complementary methods to extract and calibrate light curves
from the TESS FFIs.
First, for bright or isolated targets, we will follow \citet{montet2017} to
estimate aperture shapes and perform aperture photometry for bright sources.
Using an ensemble of sources, we will de-trend these light curves using a
modified version of \textsf{everest} \citep{luger2016, luger2017} designed to
preserve photometric signatures of rotation.
This will be achieved by fitting for the systematic effects in the light curve
using the \textsf{everest} model simultaneously with a Gaussian Process model
for the astrophysical variation.
Both \citet{aigrain2016b} and \citet{luger2016} demonstrated that this can
preserve stellar variability signals and we will use the \textsf{celerite}
algorithm \citep{dfm2017} to scale the computations to the size of TESS FFI
datasets.

@@

\section{Analysis Plan}

\section{Technical Feasibility}

\section{Expected Impact}

\section{Budget Justification}

\bibliographystyle{plainnat}
\bibliography{references}
\end{document}

>>>>

\newcommand{\Kepler}{{\it Kepler}}
\newcommand{\kepler}{\Kepler}
\newcommand{\corot}{{\it CoRoT}}
\newcommand{\Ktwo}{{\it K2}}
\newcommand{\ktwo}{\Ktwo}
\newcommand{\TESS}{{\it TESS}}
\newcommand{\LSST}{{\it LSST}}
\newcommand{\Wfirst}{{\it Wfirst}}
\newcommand{\SDSS}{{\it SDSS}}
\newcommand{\PLATO}{{\it PLATO}}
\newcommand{\Gaia}{{\it Gaia}}
\newcommand{\gaia}{{\it Gaia}}
\newcommand{\Teff}{$T_{\mathrm{eff}}$}
\newcommand{\teff}{$T_{\mathrm{eff}}$}
\newcommand{\FeH}{[Fe/H]}
\newcommand{\feh}{[Fe/H]}
\newcommand{\ie}{{\it i.e.}}
\newcommand{\eg}{{\it e.g.}}
\newcommand{\logg}{log \emph{g}}
\newcommand{\dnu}{$\Delta \nu$}
\newcommand{\numax}{$\nu_{\mathrm{max}}$}

\newcommand{\racomment}[1]{{\color{red}#1}}

\newcommand{\columbia}{1}
\newcommand{\simons}{2}
\newcommand{\ww}{3}
\newcommand{\nsf}{4}
\newcommand{\florida}{5}
\newcommand{\cca}{6}
\newcommand{\hubble}{7}
\newcommand{\princeton}{8}

@@

\author{R. Angus\altaffilmark{\columbia}
% \altaffilmark{\simons}
        J. R. Davenport\altaffilmark{\ww}
% \altaffilmark{\nsf},
        D. Busazi\altaffilmark{\florida},
        D. Foreman-Mackey\altaffilmark{\cca},
        A. Mann\altaffilmark{\columbia}
% \altaffilmark{\hubble},
        S. Oh\altaffilmark{\princeton},
        D. Kipping\altaffilmark{\columbia}
}

\altaffiltext{\columbia}{Department of Astronomy, Columbia University, New
York, NY}
\altaffiltext{\simons}{Simons Fellow, RuthAngus@gmail.com}
\altaffiltext{\ww}{Western Washington University, Bellingham, WA}
\altaffiltext{\nsf}{NSF Fellow}
\altaffiltext{\florida}{Florida Gulf Coast University, Fort Myers, FL}
\altaffiltext{\cca}{Center for Computational Astronomy, Flatiron Institute,
New York, NY}
\altaffiltext{\hubble}{Hubble Fellow}
\altaffiltext{\princeton}{Princeton University, Princeton, NJ}

@@

Galactic archaeology and exoplanet populations are two rapidly accelerating
fields of interest within astronomy.
Although seemingly unconnected, these fields are linked by a mutual
requirement for precise stellar parameters.
To galactic archaeologists, ages and elemental abundances are the most
important parameters.
Indeed, most galactic archaeology surveys target exactly these properties.
For exoplaneteers, masses and radii have historically been the most important
stellar parameters for understanding planetary systems.
With a growing number of planet hosts with precise masses and radii, attention
is turning toward other parameters, such as ages, to understand the history
and evolution of these systems.
Age is therefore a fundamental stellar parameter of great interest to at least
two large communities of astronomers.
Unfortunately however, it is a difficult attribute to measure for main sequence
F, G, K, and M stars in the field, in part because low-mass dwarfs do not move
far on the Color-Magnitude diagram (CMD) during their hydrogen burning
lifetimes.
Further, competing stellar evolution models predict different ages for the
same star.
Of the measurable properties for a large ensemble of field stars, rotation
periods contain the most information about stellar age, and provide the best
leverage for advancing our knowledge of galactic archeology as well as
exoplanet population demographics via gyrochronology.

Angular momentum is carried away though magnetically driven stellar winds,
which slows the star’s rotation over cosmic time.
This rotation-based `clock' is known as gyrochronology.
Cool spots on the star’s surface rotate in-to and out-of view, creating small
amplitude ($\pm$ 1\%) quasi-periodic changes in the stellar brightness.
While rotation periods have previously been measured from starspot-induced
flux modulations for hundreds of stars from the ground, space-based
photometric surveys have opened the door to homogeneous ensemble measures of
stellar rotation, and therefore age.
Using precise light curves available from the TESS mission, we hope to
infer ages for nearly 100,000 main sequence field stars, including thousands
of planet hosts.
The ages of the planet hosts will allow us to characterize trends in planet
properties over time.

Additionally, as rotation both influences and is influenced {\it by} stellar
magnetic fields, it is inextricably related to stellar activity.
With so many TESS targets being M dwarfs which tend to be particularly active,
understanding the magnetic behavior of these stars has never been more
important.

To enable studies of stellar ages and magnetic activity from rotation periods
with TESS, we propose to:
1. Measure accurate rotation periods for every available TESS FFI and
two-minute cadence target.
2. Produce updated gyrochronology relations based on a wider range of field
star ages.
3. Determine the origin of the mysterious rotation period bimodality
discovered with Kepler by tracing the rotation period distribution over the
whole sky, and out to further distances utilizing public Gaia data.
4. Measure the star formation history across the sky using a new Bayesian
age-dating system.
5. Further characterize the intrinsic uncertainty in gyrochronology ages using
comoving pairs identified in the first Gaia data release.

@@

The measurement of rotation periods is less sensitive to crowding and source
confusion than exoplanet transit characterization because the rotation period
is not effected by photometric dilution.
We will apply two complementary methods to extract and calibrate light curves
from the TESS FFIs.
First, for bright or isolated targets, we will follow \citet{montet2017} to
estimate aperture shapes and perform aperture photometry for bright sources.
Using an ensemble of sources, we will de-trend these light curves using a
modified version of \textsf{everest} \citep{luger2016, luger2017} designed to
preserve photometric signatures of rotation.
This will be achieved by fitting for the systematic effects in the light curve
using the \textsf{everest} model simultaneously with a Gaussian Process model
for the astrophysical variation.
Both \citet{aigrain2016} and \citet{luger2016} demonstrated that this can
preserve stellar variability signals and we will use the \textsf{celerite}
algorithm \citep{dfm2017} to scale the computations to the size of TESS FFI
datasets.

@@

\section{Expected Impact}
We will provide light curves and rotation periods for both the two-minute
cadence and FFI targets.

\section{Budget Justification}
PI Angus intends to use the budget to employ a student for 3-4 months.
In this time the student will assist in extracting light curves from the FFIs,
measuring rotation periods and building a rotation period catalog.
The student will also be involved in the scientific project of their choosing:
either the rotation period bimodality, gyrochronology of low-mass stars, or
gyrochronology of comoving pairs.

% \begin{thebibliography}
% \bibitem[Aigrain et al.(2016)]{aigrain2016} Aigrain, S., Parviainen,
%     H., \& Pope, B.~J.~S.\ 2016, \mnras, 459, 2408
% \bibitem[Angus et al.(2015)]{angus2015} Angus, R., Aigrain, S.,
% Foreman-Mackey, D., \& McQuillan, A.\ 2015, \mnras, 450, 1787
% \bibitem[Barnes(2003)]{barnes2003} Barnes, S.~A.\ 2003, \apj, 586,
% \bibitem[Barnes(2007)]{barnes2007} Barnes, S.~A.\ 2007, \apj, 669, 1167
% \bibitem[Cignoni et al.(2006)]{cignoni2006} Cignoni, M., Degl'Innocenti, S.,
% Prada Moroni, P.~G., \& Shore, S.~N.\ 2006, \aap, 459, 783 464
% \bibitem[Davenport(2017)]{davenport2017} Davenport, J.~R.~A.\ 2017, \apj, 835,
% 16
% \bibitem[Henderson et al.(2016)]{henderson2016} Henderson, C.~B.,
% Poleski, R., Penny, M., et al.\ 2016, \pasp, 128, 124401
% \bibitem[Hernandez et al.(2000)]{hernandez2000} Hernandez, X.,
% Valls-Gabaud, D., \& Gilmore, G.\ 2000, \mnras, 316, 605
% \bibitem[Luger et al.(2016)]{luger2016} Luger, R., Agol, E., Kruse, E., et
%     al.\ 2016, \aj, 152, 100
% \bibitem[Luger et al.(2017)]{luger2017} Luger, R., Kruse, E., Foreman-Mackey,
%     D., Agol, E., \& Saunders, N.\ 2017, arXiv:1702.05488
% \bibitem[Foreman-Mackey et al.(2017)]{foreman-mackey2017} Foreman-Mackey, D.,
%     Agol, E., Ambikasaran, S., \& Angus, R.\ 2017, arXiv:1703.09710
% \bibitem[McQuillan et al.(2013)]{mcquillan2013} McQuillan, A., Aigrain, S., \&
%     Mazeh, T.\ 2013, \mnras, 432, 1203
% \bibitem[McQuillan et al.(2014)]{Mcquillan2014} McQuillan, A., Mazeh, T., \&
%     Aigrain, S.\ 2014, \apjs, 211, 24
% \bibitem[Montet et al.(2017)]{montet2017} Montet, B.~T., Tovar, G., \&
%     Foreman-Mackey, D.\ 2017, arXiv:1705.07928
% \bibitem[Pace et al.(2009)]{pace2009} Pace, G., Melendez, J., Pasquini, L., et
%     al.\ 2009, \aap, 499, L9
% \bibitem[Skumanich(1972)]{skumanich1972} Skumanich, A.\ 1972, \apj, 171, 565
% \bibitem[Tremblay et al.(2014)]{tremblay2014} Tremblay, P.-E., Kalirai, J.~S.,
%     Soderblom, D.~R., Cignoni, M., \& Cummings, J.\ 2014, \apj, 791, 92
% \bibitem[Vaughan \& Preston(1980)]{vaughan1980} Vaughan, A.~H., \& Preston,
%     G.~W.\ 1980, \pasp, 92, 385
% \bibitem[Wang et al.(2016)]{wang2016} Wang, D., Hogg, D.~W., Foreman-Mackey,
%     D., \& Sch{\"o}lkopf, B.\ 2016, \pasp, 128, 094503
% \end{thebibliography}

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
